Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-14 tex/pdraft.tex
General edits.

\title{Epistemic Semantics in Guarded String Models}

@@

\begin{example}  \label{coin1} \begin{enumerate}[align=left,leftmargin=0.75em]
  \item[\apH] Amy peeks at $H$, by tipping the cup. Bob sees she's peeking, but not what she sees.
  \item[\bpH] Bob peeks at $H$.
  \item[\apT] Amy peeks at $T$.
  \item[\bpT] Bob peeks at $T$. 
  \item[\atTH] Amy secretly turns the coin from $T$ to $H$.
    She knows she turned the coin over, but not which side was face up. Bob thinks nothing happened.
  \item[\atHT] Amy secretly turns the coin from $H$ to $T$.
  \item[\atTH] Bob secretly turns the coin from $T$ to $H$.
  \item[\btHT] Bob secretly turns the coin from $H$ to $T$.
\end{enumerate}
\end{example}

@@

The events come with pre-conditions and post-conditions.  Amy can turn
the coin from heads to tails only if the coin is heads-up, so $\atHT$
has the pre-condition of the coin being heads up. Once she turns the
coin over, tails must be face-up, so $a_{10}$ has the post-condition
of the coin being tails-up. Let $h$ be the Boolean proposition that
the coin is heads up and $t$ be the Boolean proposition that the coin
is tails-up.  Then pre- and post-conditions can be described by
Boolean formulas, with $h$ being the pre-condition of $\atHT$ and
$a_{01}$ being the post-condition. This is expressed using an operator
``$:$'' (read ``and next'') that pairs Boolean formulas. The formula
$\ev h t$ describes $\atHT$ (Amy turning the coin from heads
to tails) as happening only in an $h$ state, and concluding in a
not-$h$ state.  Events don't have to change state: the event $\apT$
(Amy peeking at tails) can happen only in a $t$ state, and does
not change the state ($\ev{t}{t}$).

@@

However a coin cannot be showing both heads and tails! Currently the
precondition $h$ of $\atHT$ only says that heads must be showing, and
says nothing about the fact that tails must be face-down, indicated by
the formula $\bar t$. We will further restrict the feasible conditions
for our actions by restricting the space of valuations for our formulae.

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of atomic tests and $\phi$ is a
state constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.
Finally the lines beginning with {\tt agent} define {\em event
  alternative} relations for agents.  Each clause with an arrow has a
single event symbol on the left, and a disjunction of alternative
events on the right of the arrow. The interpretation of Amy's
alternatives for $b_1$ (Bob peeks at heads), is that when $b_1$
happens, for Amy either $b_1$ or $b_0$ (Bob peeks at tails) could be
happening. Her alternatives for $a_{01}$ and $a_{10}$ (she turns the
coin over in one state or the other) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. When Bob secretly
turns the coin over, in event $b_{10}$ or $b_{01}$, she doesn't know
anything has happened, so her alternative is the no-operation,
no-information event event $o$. Bob's event relation is symmetric.

@@

Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms over a set ${\sf{B}}$  
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
${\sf{P}}$ is the set of events $\{a_1,a_0,b_1,b_0,a_{10},a_{01},b_{10},b_{01}\}$, and
${\sf{B}}$ is $\{h,t\}$.

In the coin example, as we already saw in \ex{effect},
$\mathcal{A}^\varphi_{\mathsf B}$ is $\{h\bar{t},\bar{h}t\}$, for
which we use the shorthand $\{H,T\}$.  A guarded string over
${\sf{P}}$ and ${\sf{B}}$ is a strings of events from ${\sf{P}}$,
alternating with atoms over ${\sf{B}}$, and beginning and ending with
atoms.  In this construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$,
and $w_4=Ha_1Hb_1Ha_{10}Tb_{01}H$.

@@

\hspace{2mm}(ii) $\element{\tuple{\alpha_i,\alpha_{i+1}}}{\sem{\zeta_{ei}}^{\varphi}}$, ($0{\leq} i {\le} n$).
\end{example}

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ has the
event type $\{H\,a_1\,H\}$, and $a_0$ has the event type $\{T\,a_0\,T\}$.

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a Kripke frame $\tuple{W,R_1,...,R_n}$
consisting of a set of worlds, and a world-alternative relation for each agent
\citep{kripke1963semantical}.  An Epik specification defines an alternative relation
on bare events for each agent $a$, which we notate as ${R}_{a}$.  This should be
lifted to a relation $\hat{R}_a$ on worlds.  The basic idea is that when a world $w$
is incremented with an event $e$, in the resulting world $w \diamond e$, epistemic
alternatives for agent $a$ are of the form $w' \diamond e'$, where $w'$ is an
alternative to for $a$ in $w$, and $e'$ is and event-alternative to $e$ for
$a$.\footnote{In this it is important that the event-alternative relation for an
  agent is constant across worlds. We anticipate that the definition given here
  produces results equivalent to what is found in literature on event alternatives in
  dynamic epistemic semantics, though we have not verified this.  That literature
  primarily focuses on mapping an epistemic model for a single time and situation to
  another, and uses general first-order models, rather than guarded string
  models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and articles in
  \citet{ditmarsch2015handbook}.  Previous literature is motivated by epistemic logic
  and AI planning, rather than computable possible worlds models in natural language
  semantics.  } This needs to be implemented in a way that takes account of pre- and
post-conditions for events. For this, our approach is to refer the definition of
well-formed guarded strings.  \ex{raise1} defines a relation on worlds from a
relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over events P and primitive
  tests B, and ${R}$ be a relation on $P$. The corresponding
  relation $\hat{R}$ on $W$ holds between a guarded string
\(
\alpha_0e_0 ... e_n\alpha_{n+1} 
\)
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form
\(
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\), where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{{R}}$.
\end{example}

This requires that in an alternative world, each constituent event $e'_i$ is an
alternative to the event $e_i$ in the base world.
Compatibilities between events in the alternative world are enforced by the
requirement that the alternative world is an element of $W$, so that state and effect
formulas are enforced.

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a
set of guarded strings $\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be
the sets that are interpretations of terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of
epistemic KAT, with operations as in Figure \ref{fig:interp2}
(e.g. the binary operation $+$ is union, and the unary operation
$\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This provides a
concrete $n$-agent Kripke frame $\tuple{\sem{\bullet^*}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain of the Kripke frame
  differs from the domain of the guarded string model, because the
  former is the set of worlds, while the latter is the set of
  propositions.  } The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  It is used as a target
for natural-language interpretation in Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic KAT inside the finite state
calculus. We describe our translation from Epik terms to Fst programs
here.

@@

\begin{figure}
  \begin{tcolorbox}[size=fbox]
\footnotesize
\begin{hangingpar}
  {\tt St} \hspace{2mm} \\
  Atomic Tests such as $0110$.% The length is the number of generators.
\end{hangingpar}

\begin{hangingpar}
  {\tt UnequalStPair} \hspace{2mm} \\
  Sequence of two unequal tests such as $0110~0111$.
\end{hangingpar}

\begin{hangingpar}
\begin{verbatim}
define Wf0 ~[$ UnequalStPair];
\end{verbatim}
String that doesn't contain a non-matching test pair.
\end{hangingpar}

\begin{hangingpar}
\begin{verbatim}
define Squash St -> 0 || St _;
\end{verbatim}
Rewrite relation deleting the second of two tests.
\end{hangingpar}

@@

Translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
decorate the events $e$ with their compatible atoms. For example $a_1$ becomes
an Fst term denoting
$\{01a_101\}$. Fst has built-in operations of union (\texttt{|}),
and intersection (\texttt{\&}), which define
the sum and intersection operations in the guarded string algebra.
Fst set difference (\texttt{-}) is used to define propositional complement as the
difference between the set of worlds and the argument.

@@

KAT Kleene plus is defined in a similar way using Kleene plus in the string algebra,
with checks for equality of atoms and deletion of atoms. See Figure \ref{fstfigure1}.
KAT Kleene star is definied from KAT Kleene plus and the multiplicative identity,
which is the set of well-formed atoms \texttt{St}.

@@

\begin{tabular}{ll}
  $a$ & Relational Kleene plus in the string algebra \\
  $b$ & Constrain domain and co-domain to contain \\
      & no unmatched tests. \\
  $c$ & Reduce doubled tests to a single \\
      & test in the domain and co-domain. \\
\end{tabular}

@@

%% An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
%% bare events decorated with compatible tests on each side, $\{ 10 a_1 10
%% \}$ in this case.  This is a unit set rather than a guarded string, because elements
%% of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
%% events using {\em Kst}. The operation enforces compatibility of states, so that
%% $(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
%% Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
%% KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $R~S$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.

@@

English sentences are mapped to terms in the logical language via a
semantically interpreted multi-modal categorial grammar, consisting of
a lexicon of words, their categorial types, and interpretations in a
logical lambda language.  The grammar covers basic statives ({\em it's
  heads}), {\em that-} and {\em whether}-complements of {\em know}, predicate and
sentence negation, and predicate and sentence conjunction.  Figure
\ref{lexiconfigure} gives illustrative lexical
entries.\footnote{Category symbols use Lambek/Bar-Hillel notation for
  slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines
  with $d \backslash_D t$ on the right to give a value that combines
  with $d$ on the left to give $t$. In the semantics, lambda abstractions with
  multiple parameters are written $\lambda x~y.~e$ rather than $\lambda x.\lambda y.e$.
$d$ is the category of {\em it}.}
The grammar and semantics are
optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute
the epistemic alternative relations for those agents, rather than
individuals.  The root verb {\em know} contributes existential modal
force.  The complementizers {\em that} and {\em whether} are the heads
of their dominating clauses, and assemble an alternative relation,
modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to
express universal modal force.

>>>>

\title{Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}}
\renewcommand\footnotemark{}
%%\renewcommand\footnoterule{}

@@

\begin{example}  \label{coin1} \begin{enumerate}[align=left,leftmargin=0.75em]
  \item[\apH] Amy peeks at $H$, by tipping the cup. Bob sees she's peeking, but not what she sees.
  \item[\bpH] Bob peeks at $H$.
  \item[\apT] Amy peeks at $T$.
  \item[\bpT] Bob peeks at $T$. 
  \item[\atTH] Amy secretly turns the coin from $T$ to $H$.
    She knows she turned the coin over, but not which side was face up. Bob thinks nothing happened.
  \item[\atHT] Amy secretly turns the coin from $H$ to $T$.
  \item[\btTH] Bob secretly turns the coin from $T$ to $H$.
  \item[\btHT] Bob secretly turns the coin from $H$ to $T$.
\end{enumerate}
\end{example}

@@

The events come with pre-conditions and post-conditions.  Amy can turn
the coin from heads to tails only if the coin is heads-up, so $\atHT$
has the pre-condition of the coin being heads up. Once she turns the
coin over, tails must be face-up, so $a_{10}$ has the post-condition
of the coin being tails-up. Let $h$ be the Boolean proposition that
the coin is heads up and $t$ be the Boolean proposition that the coin
is tails-up.  Then pre- and post-conditions can be described by
Boolean formulas, with $h$ being the pre-condition of $\atHT$ and
$t$ being the post-condition. This is expressed using an operator
``$:$'' (read ``and next'') that pairs Boolean formulas. The formula
$\ev h t$ describes $\atHT$ (Amy turning the coin from heads
to tails) as happening only in an $h$ state, and concluding in a
$t$ state.  Events don't have to change state: the event $\apT$
(Amy peeking at tails) can happen only in a $t$ state, and does
not change the state ($\ev{t}{t}$).

@@

However a coin cannot be showing both heads and tails! Currently the
precondition $h$ of $\atHT$ only says that heads must be showing, and
says nothing about the fact that tails must be face-down, indicated by
the formula $\bar t$. We will further restrict the feasible conditions
for our actions by restricting the space of valuations for our formulas.

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.\footnote{Two Booleans constrained by a state formula, rather than one Boolean, are stipulated here to
 illustrate state formulas.
}  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of state primitives and $\phi$ is a
 constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.
Finally the lines beginning with {\tt agent} define {\em event
  alternative} relations for agents.  Each clause with an arrow has a
single event symbol on the left, and a disjunction of alternative
events on the right of the arrow. The interpretation of Amy's
alternatives for $b_1$ (Bob peeks at heads), is that when $b_1$
happens, for Amy either $b_1$ or $b_0$ (Bob peeks at tails) could be
happening. Her alternatives for $a_{01}$ and $a_{10}$ (she turns the
coin over in one state or the other) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. When Bob secretly
turns the coin over, in event $b_{10}$ or $b_{01}$, she doesn't know
anything has happened, so her alternative is the ``no-operation'' or
``no-information'' event event $o$. Bob's event relation is symmetric.

@@

Guarded strings over a finite alphabet ${\sf{P}}$ are like ordinary strings, but with
atoms over a set ${\sf{B}}$ alternating with the symbols from ${\sf{P}}$.  In the
algebra described by Figure \ref{epikfigure}, ${\sf{P}}$ is the set of events
$\{a_1,a_0,b_1,b_0,a_{10},a_{01},b_{10},b_{01},o\}$, and ${\sf{B}}$ is $\{h,t\}$.  As
we already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$ is
$\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$.  A guarded string
over ${\sf{P}}$ and ${\sf{B}}$ is a string of events from ${\sf{P}}$, alternating
with atoms over ${\sf{B}}$, and beginning and ending with atoms.  In this
construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$, and
$w_4=Ha_1Hb_1Ha_{10}Tb_{01}Hb_1H$.

@@

\hspace{2mm}(ii) $\element{\tuple{\alpha_i,\alpha_{i+1}}}{\sem{\zeta_{e_i}}^{\varphi}}$, ($0{\leq} i {\le} n$).
\end{example}

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ corresponds to the 
event type $\{H\,a_1\,H\}$, and $a_0$ corresponds to the event type $\{T\,a_0\,T\}$. Event
types need not be singletons.

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}
An Epik specification defines an alternative relation on bare events for each agent
$a$, which we notate as ${R}_{a}$.  This should be lifted to a relation $\hat{R}_a$
on worlds.  The basic idea is that when a world $w$ is incremented with an event $e$,
in the resulting world $w \diamond e$, epistemic alternatives for agent $a$ are of
the form $w' \diamond e'$, where $w'$ is an alternative to for $a$ in $w$, and $e'$
is and event-alternative to $e$ for $a$.\footnote{In this it is important that the
  event-alternative relation for an agent is constant across worlds. We anticipate
  that the definition given here produces results equivalent to what is found in
  literature on event alternatives in dynamic epistemic semantics, though we have not
  verified this.  That literature primarily focuses on mapping an epistemic model for
  a single time and situation to another, and uses general first-order models, rather
  than guarded string models. See \citet{baltag1999logic}, \citet{van2007dynamic},
  and articles in \citet{ditmarsch2015handbook}.  This literature is motivated by
  epistemic logic and AI planning, rather than computable possible worlds models in
  natural language semantics.  } This needs to be implemented in a way that takes
account of pre- and post-conditions for events. For this, our approach is to refer
the definition of well-formed guarded strings.  \ex{raise1} defines an epistemic
alternative relation on worlds from an alternative relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over events P and primitive
  tests B, and ${R}$ be a relation on $P$. The corresponding
  relation $\hat{R}$ on $W$ holds between a guarded string
\(
\alpha_0e_0 ... e_n\alpha_{n+1} 
\)
in $W$ and a guarded string  $q$ in $W$ iff $q$ is of the form
\(
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\), where for $0 \leq n$, $e_iRe'_i$.
\end{example}

This requires that in an alternative world, each constituent event $e'_i$ is an
alternative to the corresponding event $e_i$ in the base world.
Compatibilities between events in the alternative world are enforced by the
requirement that the alternative world is an element of $W$, so that state and effect
formulas are enforced.

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\bullet^*t \wedge \Box_a\bullet^*t + \neg\bullet^*t \wedge \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as defined
syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded strings
$\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be the sets that are interpretations of
terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of epistemic KAT, with
operations as in Figure \ref{fig:interp2} (e.g. the binary operation $+$ is union,
and the unary operation $\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This
provides a concrete $n$-agent general modal frame
$\tuple{\sem{\bullet^*}^{\epikspec}, \hat{R}_1,...,\hat{R}_n,K^{\epikspec}}$.  The
frame consists of a set of worlds, an epistemic-alternative relation for each agent,
and a set of propositions, each of which is a set of worlds.  It is used as a target
for natural-language interpretation in Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic KAT inside the finite state
calculus. We describe our translation from Epik logical terms to Fst terms
here.

@@

\begin{figure}
  \begin{tcolorbox}[size=fbox]
\footnotesize
\begin{hangingpar}
  {\tt St} \hspace{2mm} \\
  Atoms such as $0110$.% The length is the number of generators.
\end{hangingpar}

\begin{hangingpar}
  {\tt UnequalStPair} \hspace{2mm} \\
  Sequence of two unequal atoms such as $0110~0111$.
\end{hangingpar}

\begin{hangingpar}
\begin{verbatim}
define Wf0 ~[$ UnequalStPair];
\end{verbatim}
String that doesn't contain a non-matching pair of atoms.
\end{hangingpar}

\begin{hangingpar}
\begin{verbatim}
define Squash St -> 0 || St _;
\end{verbatim}
Rewrite relation deleting the second of two atoms.
\end{hangingpar}

@@

Translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
and decorate the events $e$ with their compatible atoms. For example $a_1$ becomes
an Fst term denoting
$\{01a_101\}$. Fst has built-in operations of union (\texttt{|}),
and intersection (\texttt{\&}), which are
the sum and intersection operations in the guarded string algebra as represented
in Fst.
Fst set difference (\texttt{-}) is used to define propositional complement as the
difference between the set of worlds and the argument.

@@

KAT Kleene plus is defined in a similar way using Kleene plus in the string algebra,
with checks for equality of atoms and deletion of atoms. 
KAT Kleene star is definied from KAT Kleene plus and the multiplicative identity,
which is the set of well-formed atoms \texttt{St}.

@@

\begin{tabular}{ll}
  $a$ & Relational Kleene plus in the string algebra \\
  $b$ & Constrain domain and co-domain to contain \\
      & no unmatched atoms. \\
  $c$ & Reduce doubled atoms to a single \\
      & atom in the domain and co-domain. \\
\end{tabular}

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms.  In Fst, we use the closure of the concatenation product operation on
relations to lift a relation on decorated events for an agent to the corresponding
relation on worlds.  The concatenation product $R~S$ of two relations $R$ and $S$ is
the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R y_1$, and
$x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation
  concatenation in Fst differs from relation composition ({\tt{.o.}}), and the
  closure under discussion here is the closure of the former rather than the latter.}
The epistemic alternative relation on worlds for an agent is then defined as the KAT
relation concatenation closure {\tt RelKst} of the decorated-event alternative
relation for the agent.

@@

English sentences are mapped to terms in the logical language via a semantically
interpreted multi-modal categorial grammar, consisting of a lexicon of words, their
categorial types, and interpretations in a logical lambda language.  The grammar
covers basic statives ({\em it's heads}), {\em that-} and {\em whether}-complements,
negation, and predicate and sentence
conjunction.\marginpar{Conjunction missing!}  Figure \ref{lexiconfigure} gives the
lexicon.\footnote{Category symbols use Lambek/Bar-Hillel notation for slashes, so
  that $(d \backslash t)\slash(d \backslash_D t)$ combines with $d \backslash_D t$ on
  the right to give a value that combines with $d$ on the left to give $t$. In the
  semantics, lambda abstractions with multiple parameters are written $\lambda
  x~y.~e$ rather than $\lambda x.\lambda y.e$.  $d$ is the category of expletive {\em it}.}
The grammar and semantics are optimized for a simple fragment of English concerned
with clausal complementation.  The agent names {\em Amy} and {\em Bob} contribute the
epistemic alternative relations for those agents, rather than individuals.  The root
verb {\em believe} contributes existential modal force.  The complementizers {\em that}
and {\em whether} are the heads of their dominating clauses, and assemble an
alternative relation, modal force, and proposition contributed by the complement.
These complementizers introduce the dual via two negations, in order to express
universal modal force.

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation---phrases with these top-level slashes
can only combine syntactically as arguments.
%
%For instance the category of {\em
%  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
%has category $d$.  However the phrase {\em it heads} of category $t$
%can not be formed, because $\backslash_D$ is not syntactically active
%as a function.  Instead {\em it is heads} is formed with a predicator
%{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
%Similarly {\em knows} has a category with the top-level slash
%$\slash\!_M$, and combines to form a sentence as an argument of {\em
%  that} or {\em whether}. These complementizers have a category that
%looks for the category of {\em know} on the left, after combining with
%a complement sentence on the right.
The semantic translations in the third column of Figure \ref{lexiconfigure} use
the logical language, incremented with lambda.  The body of $\lambda
x.\bullet^*h$, which is the semantic lexical entry for {\em heads}, is a
term denoting the set of all worlds where the coin is heads, expressed as the
set of all guarded strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.
%There is $\lambda x$ at the
%front because the grammar formalism enforces a strict correspondence
%between syntactic and semantic types. However, the lambda does not
%bind anything, because sentences such as {\em it isn't heads} have an
%expletive subject.
The body of $\lambda p.\lambda R. {\Diamond}_R p$, which is the semantic lexical
entry of {\em believes}, is an term denoting the pre-image of the world-alternative
relation contributed by the subject.  This is not the right semantics for {\em Amy
  believes that it's heads}, because it has an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the
complementizer {\em that}, which introduces the dual.
